---
title: Stochastic simulation of evolution of antibiotic resistance in the face of
  anorexia
output: bookdown::html_document2
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      dev=c('png','tiff'),
                      fig.path='figures/')
```

## Introduction

I am considering the following very simple model of resource-dependent competition between two parasite strains, $P_1$ and $P_2$:
\begin{align}
\frac{dR}{dt} &= \theta - \chi R - \frac{\alpha_1 R}{\eta_1 + R}P_1 - \frac{\alpha_2 R}{\eta_2 + R}P_2 \\
\frac{dP_1}{dt} &= \epsilon_1 \frac{\alpha_1 R}{\eta_1 + R}P_1 - \delta_1 P_1 - \mu P_1 + \mu P_2 \\
\frac{dP_1}{dt} &= \epsilon_2 \frac{\alpha_2 R}{\eta_2 + R}P_2 - \delta_2 P_2 - \mu P_2 + \mu P_1
\end{align}

In this model, resources come into the system through "ingestion," $\theta$, and are utilized by, or lost to, the host at the per-capita rate $\chi$. 
The two parasite strains each have a Type II functional response with separate acquisition rates and half-saturation constants, $\alpha_i$ and $\eta_i$. 
Each parasite converts resources into new parasites with efficiency $\epsilon_i$, and are killed at the rate $\delta_i$.
Note that we could approximate the effect of the immune system by changing the mortality rate from a linear function to a quadratic function (e.g., if you assume that the immune response dynamics are something like $dI/dt = a (P_1+P_2) - m I$, and you assume that this attains a quasi-equilibrium [e.g., a separation of timescales argument], then $\hat{I} = a (P_2+P_2)/m$ and the parasite mortality rate would be $\delta_1 P_1 I = \delta_1 a (P_1^2 + P_1 P_2)/m$).
Finally, there is mutation between the two strains at a rate $\mu$. 

The existence of mutation implies that a deterministic model will always have both strains present.
However, if you analyze a model where $\mu = 0$, then coexistence is impossible, and the species that will win is the one that suppresses resources the most.
The resource abundance at equilibrium for either strain alone is $\hat{R} = \eta_i \delta_i/(\epsilon_i \alpha_i - \delta_i)$, so if 
\begin{equation}
\frac{\eta_1 \delta_1}{\epsilon_1 \alpha_1 - \delta_1} > \frac{\eta_2 \delta_2}{\epsilon_2 \alpha_2 - \delta_2}
\end{equation}
then strain 1 will outcompete strain 2, and vice versa if the inequality is reversed.
 
Additionally, analysis shows that, in the absence of mutation, the single strain models always go to a stable equilibrium so long as
\begin{equation}
\frac{\theta}{\chi} > \frac{\eta_i \delta_i}{(\epsilon_i \alpha_i - \delta_i)},
\end{equation}
where $\theta/\chi$ is the resource equilibrium in the absence of infection. 
Thus, if hosts drop $\theta$ enough in the presence of infection, violating the above inequality, then no parasite strains can persist at all.
Another way to think about this is that, even without competition (or, importantly, when we add drugs that could alter the competitive outcome), if the following equalities hold
\begin{align}
\frac{\theta}{\chi} &> \frac{\eta_1 \delta_1}{(\epsilon_1 \alpha_1 - \delta_1)} \\
\frac{\theta}{\chi} &< \frac{\eta_2 \delta_2}{(\epsilon_2 \alpha_2 - \delta_2)} 
\end{align}
then strain 1 will win and strain 2 will be outcompeted.

As a preliminary analysis, I will do stochastic simulations of the above model assuming that parasite strain 1 can persist and outcompete strain 2 in the absence of mutation.

```{r sim1, echo=FALSE}
stochastic_strain_competition <- function(tmax, tstep, pars, y0, seed) {
  set.seed(seed)
  
  theta <- pars["theta"]
  chi <- pars["chi"]
  alpha1 <- pars["alpha1"]
  alpha2 <- pars["alpha2"]
  eta1 <- pars["eta1"]
  eta2 <- pars["eta2"]
  epsilon1 <- pars["epsilon1"]
  epsilon2 <- pars["epsilon2"]
  delta1 <- pars["delta1"]
  delta2 <- pars["delta2"]
  mu <- pars["mu"]
  
  ## initialize the time at t = 0
  t <- 0
  
  R <- y0["R"]
  P1 <- y0["P1"]
  P2 <- y0["P2"]
  
  ## initialize a dataframe to store the time and population size
  sim <- data.frame(time=seq(0,tmax,tstep), 
                    R=rep(0, length(seq(0,tmax,tstep))),
                    P1=rep(0, length(seq(0,tmax,tstep))),
                    P2=rep(0, length(seq(0,tmax,tstep))))
                    
  sim[1,2:4] <- c(R,P1,P2)
  
  ## start the algorithm
  while (t < tmax & (P1+P2) > 0) {
    ## compute rates
    rates <- c(theta, ## add resource = 1
               chi*R, ## host resource use = 2
               alpha1*R/(eta1+R)*P1, ## strain 1 resource consumption = 3
               alpha2*R/(eta2+R)*P2, ## strain 2 resource consumption = 4
               epsilon1*alpha1*R/(eta1+R)*P1, ## strain 1 replication = 5
               epsilon2*alpha2*R/(eta2+R)*P2, ## strain 2 replication = 6
               delta1*P1, ## strain 1 death = 7
               delta2*P2, ## strain 2 death = 8
               mu*P1, ## strain 1 mutating to strain 2 = 9
               mu*P2) ## strain 2 mutating to strain 1 = 10
   
    ## what time does the event happen?
    dt <- rexp(1, rate=sum(rates))
    
    ## if these are not equal, don't record anything
    if(max(which(t >= seq(0,tmax,tstep)))!=max(which(t+dt > seq(0,tmax,tstep))))
      sim[max(which(t+dt > seq(0,tmax,tstep))), 2:4] <- c(R, P1, P2)
    
    ## update t 
    t <- t + dt
    
    ## "wheel of fortune"
    wheel <- cumsum(rates)/sum(rates)
    
    ## which event happens? Draw a random uniform to determine
    rand <- runif(1)
    event <- 1 + sum(rand > wheel)
    if (event==1) R <- R + 1
    else if (event%in%2:4) R <- R - 1
    else if (event==5) P1 <- P1 + 1
    else if (event==6) P2 <- P2 + 1
    else if (event==7) P1 <- P1 - 1
    else if (event==8) P2 <- P2 - 1
    else if (event==9) { P1 <- P1 -1; P2 <- P2 + 1}
    else {P1 <- P1 + 1; P2 <- P2 - 1}
      
  }
  return(sim)
}

## Deterministic version of the model
deterministic_strain_competition <- function(t, y0, pars) {
  theta <- pars["theta"]
  chi <- pars["chi"]
  alpha1 <- pars["alpha1"]
  alpha2 <- pars["alpha2"]
  eta1 <- pars["eta1"]
  eta2 <- pars["eta2"]
  epsilon1 <- pars["epsilon1"]
  epsilon2 <- pars["epsilon2"]
  delta1 <- pars["delta1"]
  delta2 <- pars["delta2"]
  mu <- pars["mu"]
  
  R <- y0["R"]
  P1 <- y0["P1"]
  P2 <- y0["P2"]
  b <- pars["b"]
  bs <- pars["bs"]
  d <- pars["d"]
  ds <- pars["ds"]
  
  dRdt <- theta - chi*R - alpha1*R/(eta1+R)*P1 - alpha1*R/(eta1+R)*P2
  dP1dt <- epsilon1*alpha1*R/(eta1+R)*P1 - delta1*P1 - mu*P1 + mu*P2
  dP2dt <- epsilon2*alpha2*R/(eta2+R)*P2 - delta2*P2 - mu*P2 + mu*P1
  
  return(list(c(dRdt,dP1dt,dP2dt)))
}

library(deSolve)
library(parallel)
library(magrittr)

```

```{r fig1, echo=FALSE, fig.height=4, fig.width=6, fig.cap="Dynamics of resources and each pathogen strain when resource competition is fairly similar between the two strains. For this set of parameters, the strain 1-only resource equilibrium is 52.6, and the strain 2-only resoure equilibrim is 58.8. In the absence of parasites, the equilibrium is 100. The mutation rate is fairly high, 0.01."}

## choose parameters appropriately so that strain 1 can persist and outcompete strain 2
pars <- c(theta=10, chi=0.1, alpha1=1, alpha2=0.9, eta1=1000, eta2=1000, epsilon1=20, epsilon2=20, delta1=1, delta2=1, mu=0.01)
# ## resource abundance in the absence of parasites
# pars["theta"]/pars["chi"]
# ## resource abundance with strain 1
# R1 <- pars["eta1"]*pars["delta1"]/(pars["epsilon1"]*pars["alpha1"]-pars["delta1"])
# ## strain 1 outcompetes strain 2 if this is smaller than the above
# R2 <- pars["eta2"]*pars["delta2"]/(pars["epsilon2"]*pars["alpha2"]-pars["delta2"])
# ## strain 1 abundance at equilibrium in the absence of mutation
# -(((pars["eta1"] + R1)*(pars["chi"]*R1 - pars["theta"]))/(pars["alpha1"]*R1))

## Set the initial conditions
y0 <- c(R = 100, P1 = 10, P2 = 0)

## Do some stochastic simulations
set.seed(1234)
seeds <- floor(runif(24, 1, 1e5))
mclapply(seeds,
         function(s) stochastic_strain_competition(100, tstep=0.1, pars=pars, y0=y0, seed=s),
         mc.cores=4) -> stoch_sims

par(mfrow=c(1,3))
for (j in 2:4) {
  plot.new()
  plot.window(xlim=range(stoch_sims[[1]][,1]), ylim=c(0, lapply(stoch_sims, function(s) max(s[,j])) %>% unlist %>% max))
  axis(1); axis(2); box('plot')
  for (i in 1:length(stoch_sims)) lines(stoch_sims[[i]][,1], stoch_sims[[i]][,j], col=gray(0.5))
  lines(stoch_sims[[1]][,1], lapply(stoch_sims, function(s) s[,j]) %>% do.call(cbind.data.frame, .) %>% apply(., 1, mean), col=1, lwd=2)
  if(j==2) mtext(side=2, line=2, "Resource abundance")
  if(j==3) mtext(side=2, line=2, "Drug-susceptible abundance")
  if(j==4) mtext(side=2, line=2, "Drug-resistant abundance")
}


```

```{r fig2, echo=FALSE, fig.height=4, fig.width=6, fig.cap="Dynamics of resources and each pathogen strain when resource competition is more biased towards strain 1. For this set of parameters, the strain 1-only resource equilibrium is 52.6, and the strain 2-only resoure equilibrim is 76.9. In the absence of parasites, the equilibrium is 100. The mutation rate is fairly high, 0.01. This change makes the mutant strain much rarer, as it tends to be outcompeted fairly quickly."}

## choose parameters appropriately so that strain 1 can persist and outcompete strain 2
pars <- c(theta=10, chi=0.1, alpha1=1, alpha2=0.7, eta1=1000, eta2=1000, epsilon1=20, epsilon2=20, delta1=1, delta2=1, mu=0.01)
# ## resource abundance in the absence of parasites
# pars["theta"]/pars["chi"]
# ## resource abundance with strain 1
# R1 <- pars["eta1"]*pars["delta1"]/(pars["epsilon1"]*pars["alpha1"]-pars["delta1"])
# ## strain 1 outcompetes strain 2 if this is smaller than the above
# R2 <- pars["eta2"]*pars["delta2"]/(pars["epsilon2"]*pars["alpha2"]-pars["delta2"])
# ## strain 1 abundance at equilibrium in the absence of mutation
# -(((pars["eta1"] + R1)*(pars["chi"]*R1 - pars["theta"]))/(pars["alpha1"]*R1))

## Set the initial conditions
y0 <- c(R = 100, P1 = 10, P2 = 0)

## Do some stochastic simulations
set.seed(1234)
seeds <- floor(runif(24, 1, 1e5))
mclapply(seeds,
         function(s) stochastic_strain_competition(100, tstep=0.1, pars=pars, y0=y0, seed=s),
         mc.cores=4) -> stoch_sims

par(mfrow=c(1,3))
for (j in 2:4) {
  plot.new()
  plot.window(xlim=range(stoch_sims[[1]][,1]), ylim=c(0, lapply(stoch_sims, function(s) max(s[,j])) %>% unlist %>% max))
  axis(1); axis(2); box('plot')
  for (i in 1:length(stoch_sims)) lines(stoch_sims[[i]][,1], stoch_sims[[i]][,j], col=gray(0.5))
  lines(stoch_sims[[1]][,1], lapply(stoch_sims, function(s) s[,j]) %>% do.call(cbind.data.frame, .) %>% apply(., 1, mean), col=1, lwd=2)
  if(j==2) mtext(side=2, line=2, "Resource abundance")
  if(j==3) mtext(side=2, line=2, "Drug-susceptible abundance")
  if(j==4) mtext(side=2, line=2, "Drug-resistant abundance")
}


```

```{r fig3, echo=FALSE, fig.height=4, fig.width=6, fig.cap="This is the same set of parameters as in the preceding simulation, except that the mutation rate is much lower, 0.001. Now the mutant essentially cannot persist."}

## choose parameters appropriately so that strain 1 can persist and outcompete strain 2
pars <- c(theta=10, chi=0.1, alpha1=1, alpha2=0.7, eta1=1000, eta2=1000, epsilon1=20, epsilon2=20, delta1=1, delta2=1, mu=0.001)
# ## resource abundance in the absence of parasites
# pars["theta"]/pars["chi"]
# ## resource abundance with strain 1
# R1 <- pars["eta1"]*pars["delta1"]/(pars["epsilon1"]*pars["alpha1"]-pars["delta1"])
# ## strain 1 outcompetes strain 2 if this is smaller than the above
# R2 <- pars["eta2"]*pars["delta2"]/(pars["epsilon2"]*pars["alpha2"]-pars["delta2"])
# ## strain 1 abundance at equilibrium in the absence of mutation
# -(((pars["eta1"] + R1)*(pars["chi"]*R1 - pars["theta"]))/(pars["alpha1"]*R1))

## Set the initial conditions
y0 <- c(R = 100, P1 = 10, P2 = 0)

## Do some stochastic simulations
set.seed(1234)
seeds <- floor(runif(24, 1, 1e5))
mclapply(seeds,
         function(s) stochastic_strain_competition(100, tstep=0.1, pars=pars, y0=y0, seed=s),
         mc.cores=12) -> stoch_sims

par(mfrow=c(1,3))
for (j in 2:4) {
  plot.new()
  plot.window(xlim=range(stoch_sims[[1]][,1]), ylim=c(0, lapply(stoch_sims, function(s) max(s[,j])) %>% unlist %>% max))
  axis(1); axis(2); box('plot')
  for (i in 1:length(stoch_sims)) lines(stoch_sims[[i]][,1], stoch_sims[[i]][,j], col=gray(0.5))
  lines(stoch_sims[[1]][,1], lapply(stoch_sims, function(s) s[,j]) %>% do.call(cbind.data.frame, .) %>% apply(., 1, mean), col=1, lwd=2)
  if(j==2) mtext(side=2, line=2, "Resource abundance")
  if(j==3) mtext(side=2, line=2, "Drug-susceptible abundance")
  if(j==4) mtext(side=2, line=2, "Drug-resistant abundance")
}


```

How does anorexia affect these results? We can explore that by varying the value of $\theta$ and asking how that affects the abundances of both strains at equilibrium.

```{r fig4, echo=FALSE, fig.height=8, fig.width=6, fig.cap="Mean resource and pathogen strain abundances across stochastic simulations for different levels of host resource ingestion."}

## Vary theta and record the mean trajectories only
## Let theta = 10 be the "normal" feeding level, so theta=8, 9 represent moderate and mild anorexia
Rresults <- P1results <- P2results <- c()
for (th in 8:10) {
  pars <- c(theta=th, chi=0.1, alpha1=1, alpha2=0.7, eta1=1000, eta2=1000, epsilon1=20, epsilon2=20, delta1=1, delta2=1, mu=0.001)
  set.seed(1234)
  seeds <- floor(runif(24, 1, 1e5))
  mclapply(seeds,
           function(s) stochastic_strain_competition(100, tstep=0.1, pars=pars, y0=c(R=100, P1=10, P2=0), seed=s),
           mc.cores=12) -> stoch_sims
  Rresults <- cbind(Rresults,
                   lapply(stoch_sims, function(s) s[,2]) %>% do.call(cbind.data.frame, .) %>% apply(., 1, mean))
  P1results <- cbind(P1results,
                   lapply(stoch_sims, function(s) s[,3]) %>% do.call(cbind.data.frame, .) %>% apply(., 1, mean))
  P2results <- cbind(P2results,
                   lapply(stoch_sims, function(s) s[,4]) %>% do.call(cbind.data.frame, .) %>% apply(., 1, mean))
  
}

par(mfrow=c(3,1), mar=c(4,4,0,0), oma=rep(0.5,4))
plot.new()
plot.window(xlim=c(0,100), ylim=range(Rresults[,1:3]))
axis(1); axis(2); box('plot')
for(i in 1:3) lines(seq(0,100,0.1), Rresults[,i], col=i, lwd=2)
mtext(side=1, line=2, "Time")
mtext(side=2, line=2, "Resource abundance")

plot.new()
plot.window(xlim=c(0,100), ylim=range(P1results[,1:3]))
axis(1); axis(2); box('plot')
for(i in 1:3) lines(seq(0,100,0.1), P1results[,i], col=i, lwd=2)
mtext(side=1, line=2, "Time")
mtext(side=2, line=2, "Drug-susceptible abundance")

plot.new()
plot.window(xlim=c(0,100), ylim=range(P2results[,1:3]))
axis(1); axis(2); box('plot')
for(i in 1:3) lines(seq(0,100,0.1), P2results[,i], col=i, lwd=2)
mtext(side=1, line=2, "Time")
mtext(side=2, line=2, "Drug-resistant abundance")


```
  
Key thing to note here: increasing ingestion increases the abundance of strain 1, but not strain 2 in the absence of drugs. 
This is the key thing to keep in mind... anorexia will reduce the likelihood of emergence by reducing the abundance of the drug-susceptible strain.

What if you allow for drugs?
Here I am letting the parameter $\Delta$ capture the effect of drugs on strain 1 only - I will assume for simplicity that strain 2 is completely resistant to all doses of drugs.
We also have to assume that drugs are not introduced until some time in the future, since they won't be there at the beginning of infection.
Here I will assume that drugs start at $t = 20$.
\begin{align}
\frac{dR}{dt} &= \theta - \chi R - \frac{\alpha_1 R}{\eta_1 + R}P_1 - \frac{\alpha_2 R}{\eta_2 + R}P_2 \\
\frac{dP_1}{dt} &= \epsilon_1 \frac{\alpha_1 R}{\eta_1 + R}P_1 - \delta_1 P_1 - \mu P_1 + \mu P_2 -\Delta \\
\frac{dP_1}{dt} &= \epsilon_2 \frac{\alpha_2 R}{\eta_2 + R}P_2 - \delta_2 P_2 - \mu P_2 + \mu P_1
\end{align}

When ingestion is high, I find that increasing drug pressure increases the likelihood of resistance emerging (Fig. \@ref(fig:highIngestFig1)). 
This might actually strike as a bit surprising, since the argument for high-dose drugs is that they crush drug-susceptible populations quickly, thereby limiting mutation supply.

```{r, echo=FALSE, eval=FALSE}
stochastic_strain_competition_w_drugs <- function(tmax, tstep, pars, y0, seed) {
  set.seed(seed)
  
  theta <- pars["theta"]
  chi <- pars["chi"]
  alpha1 <- pars["alpha1"]
  alpha2 <- pars["alpha2"]
  eta1 <- pars["eta1"]
  eta2 <- pars["eta2"]
  epsilon1 <- pars["epsilon1"]
  epsilon2 <- pars["epsilon2"]
  delta1 <- pars["delta1"]
  delta2 <- pars["delta2"]
  mu <- pars["mu"]
  D <- pars["D"]
  tDrug <- pars["tDrug"]
  
  ## initialize the time at t = 0
  t <- 0
  
  R <- y0["R"]
  P1 <- y0["P1"]
  P2 <- y0["P2"]
  
  ## initialize a dataframe to store the time and population size
  sim <- array(NA, dim=c(1e5, 4))
  sim[1,] <- c(t,R,P1,P2)
  i <- 2
  ## start the algorithm
  while (t < tmax) {
    ## compute rates
    rates <- c(theta, ## add resource = 1
               chi*R, ## host resource use = 2
               alpha1*R/(eta1+R)*P1, ## strain 1 resource consumption = 3
               alpha2*R/(eta2+R)*P2, ## strain 2 resource consumption = 4
               epsilon1*alpha1*R/(eta1+R)*P1, ## strain 1 replication = 5
               epsilon2*alpha2*R/(eta2+R)*P2, ## strain 2 replication = 6
               ifelse(P1 > 0, ifelse(t > tDrug, delta1*P1 + D, delta1*P1), 0), ## strain 1 death = 7
               delta2*P2, ## strain 2 death = 8
               mu*P1, ## strain 1 mutating to strain 2 = 9
               mu*P2) ## strain 2 mutating to strain 1 = 10
   
    ## what time does the event happen?
    dt <- rexp(1, rate=sum(rates))
    
    ## update t 
    t <- t + dt
    
    ## "wheel of fortune"
    wheel <- cumsum(rates)/sum(rates)
    
    ## which event happens? Draw a random uniform to determine
    rand <- runif(1)
    event <- 1 + sum(rand > wheel)
    if (event==1) R <- R + 1
    else if (event%in%2:4) R <- R - 1
    else if (event==5) P1 <- P1 + 1
    else if (event==6) P2 <- P2 + 1
    else if (event==7) P1 <- P1 - 1
    else if (event==8) P2 <- P2 - 1
    else if (event==9) { P1 <- P1 -1; P2 <- P2 + 1}
    else {P1 <- P1 + 1; P2 <- P2 - 1}
    
    sim[i,] <- c(t, R, P1, P2) 
    i <- i + 1
  }
  sim <- sim[1:(i-1),]
  ## Clean up: we're recording **way** too much information here - only record every tstep time steps
  lapply(seq(0,tmax,tstep), function(t) c(t,sim[((sim[,1] > t) %>% which %>% min)-1,2:4])) %>% do.call(rbind.data.frame,.) -> sim2
  colnames(sim2) <- c("t","R","P1","P2")
  return(sim2)
}

seeds <- readRDS("seeds.RDS")
## simulate with 
lapply(seq(0,5,0.1), 
       function(drug) {
         print(drug)
         pars <- c(theta=10, chi=0.1, alpha1=0.1, alpha2=0.09, eta1=1500, eta2=1500, epsilon1=20, epsilon2=20, delta1=0.1, delta2=0.1, mu=0.001, D=drug, tDrug=150)
         mclapply(seeds, 
                  function(s) stochastic_strain_competition_w_drugs(500, tstep=1, pars=pars, y0=c(R=unname(pars["theta"]/pars["chi"]), P1=10, P2=0), seed=s),
                  mc.cores=4)
       }) -> out_drug_no_anorexia_hi

lapply(seq(0,5,0.1), 
       function(drug) {
         print(drug)
         pars <- c(theta=10, chi=0.1, alpha1=0.1, alpha2=0.09, eta1=1500, eta2=1500, epsilon1=20, epsilon2=20, delta1=0.1, delta2=0.1, mu=0.0001, D=drug, tDrug=150)
         mclapply(seeds, 
                  function(s) stochastic_strain_competition_w_drugs(500, tstep=1, pars=pars, y0=c(R=unname(pars["theta"]/pars["chi"]), P1=10, P2=0), seed=s),
                  mc.cores=4)
       }) -> out_drug_no_anorexia_lo

saveRDS(out_drug_no_anorexia_hi, file="out_drug_no_anorexia_hi.RDS")
saveRDS(out_drug_no_anorexia_lo, file="out_drug_no_anorexia_lo.RDS")
```

```{r highIngestFig1, echo=FALSE, fig.height=4, fig.width=6, fig.cap="Probability of drug-resistance emerging when ingestion is high."}

out10 <- readRDS("out_drug_no_anorexia_hi.RDS")

## measure the probability of emergence as the probability that the abundance of the resistant strain over the final 10 time points is higher than the expected abundance based on the drug-free situation
cutoff <- lapply(out10[[1]], function(o) tail(o[,4],10) %>% mean) %>% unlist %>% mean(.,na.rm=TRUE)

lapply(1:length(out10),
       function(i) (lapply(out10[[i]], function(o) (ifelse((!is.na(tail(o[,4],1))),tail(o[,4],10) %>% mean,0) > 2*cutoff)) %>% unlist %>% sum)/length(out10[[i]])) %>% unlist -> emergeProb10

par(mar=c(3,5,0.5,0.5))
plot(seq(0,5,0.1), emergeProb10, pch=21, bg=1, cex=1.5, xaxt='n', ylab="Probability of resistance", cex.lab=1.5, cex.axis=1.65)
axis(1,tick=TRUE,labels=FALSE)
mtext(side=1, line=1.5, "Drug dose", cex=1.65)
```

In Fig. \@ref(fig:highIngestFig2), you can see that the "crushing" effect of the drugs on the drug-susceptible strain is still evident, as the time between drug onset and susceptible extinction decreases with increasing drug dose.
Even increasing the dose greatly (from 5 to 20) doesn't change the expected probability of resistance emerging, even though it does continue to decrease the time until susceptible extinction.
There is, of course, a straightforward explanation for this: if resistance has emerged, whether dose is 5 or 20 is irrelevant - it's enough to give a big fitness advantage to the drug susceptible strain and thus to facilitate the evolution of resistance.

```{r highIngestFig2, echo=FALSE, fig.height=4, fig.width=6, fig.cap="Probability of drug-resistance emerging when ingestion is high."}

## Time to extinction:
lapply(1:length(out10), 
       function(i) lapply(out10[[i]], 
                          function(o) ifelse(tail(o[,1],1) > 150, 
                                             c(o[(o[,1] > 150) %>% which %>% min,1], ifelse(any(o[,3]==0), 
                                                                                            o[which(o[,3]==0) %>% min,1], 
                                                                                            350)) %>% diff,
                                             NA)
                          ) %>% unlist %>% mean(.,na.rm=TRUE)
) %>% unlist -> timeToExtinction
    
par(mar=c(3,5,0.5,0.5))
plot(seq(0,5,0.1), timeToExtinction, pch=21, bg=1, cex=1.5, xaxt='n', ylab="Time until susceptible extinction", cex.lab=1.5, cex.axis=1.65)
axis(1,tick=TRUE,labels=FALSE)
mtext(side=1, line=1.5, "Drug dose", cex=1.65)               
    
## Even increasing dose to 20 doesn't change the fundamental result
# pars <- c(theta=10, chi=0.1, alpha1=0.1, alpha2=0.09, eta1=1500, eta2=1500, epsilon1=20, epsilon2=20, delta1=0.1, delta2=0.1, mu=0.001, D=40, tDrug=150)
# mclapply(seeds, 
#          function(s) stochastic_strain_competition_w_drugs(500, tstep=0.1, pars=pars, y0=c(R=unname(pars["theta"]/pars["chi"]), P1=10, P2=0), seed=s),
#          mc.cores=12) -> ou
# ## Proability of emergence doesn't change
# 1-(lapply(ou, function(o) tail(o[,4],1)) %>% unlist %>% is.na %>% sum)/length(ou)
# ## But mean time to clearance decreases some more
# (lapply(ou, function(o) o[which(o[,3]==0) %>% min,1]) %>% unlist %>% mean(.,na.rm=T))-150
```

To test that hypothesis, I can look at the probability of resistance emerging with dose when there were no resistant strains around at the time of drug treatment starting (Fig. \@ref(fig:highIngestFig3)).

```{r highIngestFig3, echo=FALSE, fig.width=6, fig.height=5, fig.cap="Comparing the amount of drug resistance present at the beginning and the end of the simulations across multiple drug doses. What is evident here is that, at low doses, even if abundance is low at the beginning of drugs, there is a chance of emergence, although it is contained; at high doses, though, if resistance isn't there at the beginning, it is very unlikely to evolve."}
par(mfrow=c(2,3), mar=c(3,3,0.5,0.5), oma=c(3,3,0,0))
lapply(out10[[1]], function(o) c(o[151,4],o[501,4])) %>% do.call(rbind.data.frame,.) -> x
x[is.na(x)] <- 0
plot(x[,1], x[,2], xlab="", ylab="")
legend(x='topleft', "Dose = 0", bty='n')

lapply(out10[[5]], function(o) c(o[151,4],o[501,4])) %>% do.call(rbind.data.frame,.) -> x
x[is.na(x)] <- 0
plot(x[,1], x[,2], xlab="", ylab="")
legend(x='topleft', "Dose = 1", bty='n')

lapply(out10[[9]], function(o) c(o[151,4],o[501,4])) %>% do.call(rbind.data.frame,.) -> x
x[is.na(x)] <- 0
plot(x[,1], x[,2], xlab="", ylab="")
legend(x='topleft', "Dose = 2", bty='n')

lapply(out10[[13]], function(o) c(o[151,4],o[501,4])) %>% do.call(rbind.data.frame,.) -> x
x[is.na(x)] <- 0
plot(x[,1], x[,2], xlab="", ylab="")
legend(x='topleft', "Dose = 3", bty='n')

lapply(out10[[17]], function(o) c(o[151,4],o[501,4])) %>% do.call(rbind.data.frame,.) -> x
x[is.na(x)] <- 0
plot(x[,1], x[,2], xlab="", ylab="")
legend(x='topleft', "Dose = 4", bty='n')

lapply(out10[[21]], function(o) c(o[151,4],o[501,4])) %>% do.call(rbind.data.frame,.) -> x
x[is.na(x)] <- 0
plot(x[,1], x[,2], xlab="", ylab="")
legend(x='topleft', "Dose = 5", bty='n')
mtext(side=1, outer=T, "Resistance at start of drugs")
mtext(side=2, outer=T, "Resistance at end")

```

The final confirmation comes from reducing the mutation rate and seeing whether that changes the overall pattern, making it less likely for resistance to emerge at higher doses (Fig. \@ref(fig:highIngestFig4)).

```{r highIngestFig4, echo=FALSE, fig.height=4, fig.width=6, fig.cap="Probability of drug-resistance emerging when ingestion is high but the mutation rate is much lower (0.0001). Here you can see a strong unimodal relationship between the probability of resistance emerging and drug dose, as expected where competition constrains the evolution of resistance at low doses."}

out10 <- readRDS("out_drug_no_anorexia_lo.RDS")

## measure the probability of emergence as the probability that the abundance of the resistant strain over the final 10 time points is higher than the expected abundance based on the drug-free situation
cutoff <- lapply(out10[[1]], function(o) tail(o[,4],10) %>% mean) %>% unlist %>% mean(.,na.rm=TRUE)

lapply(1:length(out10),
       function(i) (lapply(out10[[i]], function(o) (ifelse((!is.na(tail(o[,4],1))),tail(o[,4],10) %>% mean,0) > 2*cutoff)) %>% unlist %>% sum)/length(out10[[i]])) %>% unlist -> emergeProb10

par(mar=c(3,5,0.5,0.5))
plot(seq(0,5,0.1), emergeProb10, pch=21, bg=1, cex=1.5, xaxt='n', ylab="Probability of resistance", cex.lab=1.5, cex.axis=1.65)
axis(1,tick=TRUE,labels=FALSE)
mtext(side=1, line=1.5, "Drug dose", cex=1.65)

D <- seq(0,5,0.1)
D2 <- D^2
lines(D, predict.lm(lm(emergeProb10~D+D2), newdata=data.frame(D=D, D2=D^2)), col=2)


```

All of the preceding, however, doesn't consider the effect of anorexia.
As discussed above, if the host drops ingestion severely enough, the drug-resistant strain will not be able to emerge because the host's resource ingestion will be too low.
You can see that very clearly in the simulations below.
With this set of parameter values, in the absence of drugs or competition, the resource level that the drug-resistant strain requires (its $R^*$ value) is 88.2, whereas the drug-susceptible $R^*$ is 78.9. 
If I set the host resource ingestion to $\theta=8$, then the resource equilibrium in the absence of any pathogens is $\theta/\chi=8/0.1=80$, which is too low for the drug-resistant strain. 
As expected, drug-resistance never emerges, regardless of the mutation rate (Fig. \@ref(fig:lowIngestFig)).

```{r, echo=FALSE, eval=FALSE}
lapply(seq(0,5,0.1), 
       function(drug) {
         print(drug)
         pars <- c(theta=8, chi=0.1, alpha1=0.1, alpha2=0.09, eta1=1500, eta2=1500, epsilon1=20, epsilon2=20, delta1=0.1, delta2=0.1, mu=0.001, D=drug, tDrug=150)
         mclapply(seeds, 
                  function(s) stochastic_strain_competition_w_drugs(500, tstep=1, pars=pars, y0=c(R=unname(pars["theta"]/pars["chi"]), P1=10, P2=0), seed=s),
                  mc.cores=4)
       }) -> out_drug_anorexia_hi

lapply(seq(0,5,0.1), 
       function(drug) {
         print(drug)
         pars <- c(theta=8, chi=0.1, alpha1=0.1, alpha2=0.09, eta1=1500, eta2=1500, epsilon1=20, epsilon2=20, delta1=0.1, delta2=0.1, mu=0.0001, D=drug, tDrug=150)
         mclapply(seeds, 
                  function(s) stochastic_strain_competition_w_drugs(500, tstep=1, pars=pars, y0=c(R=unname(pars["theta"]/pars["chi"]), P1=10, P2=0), seed=s),
                  mc.cores=4)
       }) -> out_drug_anorexia_lo
saveRDS(out_drug_anorexia_hi, file="out_drug_anorexia_hi.RDS")
saveRDS(out_drug_anorexia_lo, file="out_drug_anorexia_lo.RDS")


```
```{r lowIngestFig, echo=FALSE, fig.height=4, fig.width=6, fig.cap="Probability of drug-resistance emerging when ingestion is low. Regardless of mutation rate, drug resistance does not emerge because the host ingestion is such that there are not enough resources under any circumstance to sustain the drug resistant strain."}

out_drug_anorexia_hi <- readRDS("out_drug_anorexia_hi.RDS")
out_drug_anorexia_lo <- readRDS("out_drug_anorexia_lo.RDS")

## measure the probability of emergence as the probability that the abundance of the resistant strain over the final 10 time points is higher than the expected abundance based on the drug-free situation
lapply(1:length(out_drug_anorexia_hi),
       function(i) (lapply(out_drug_anorexia_hi[[i]], function(o) (ifelse((!is.na(tail(o[,4],1))),tail(o[,4],10) %>% mean,0) > 20)) %>% unlist %>% sum)/length(out_drug_anorexia_hi[[i]])) %>% unlist -> emergeProb1

lapply(1:length(out_drug_anorexia_lo),
       function(i) (lapply(out_drug_anorexia_lo[[i]], function(o) (ifelse((!is.na(tail(o[,4],1))),tail(o[,4],10) %>% mean,0) > 20)) %>% unlist %>% sum)/length(out_drug_anorexia_lo[[i]])) %>% unlist -> emergeProb2

par(mar=c(3,5,0.5,0.5))
plot(seq(0,5,0.1), emergeProb1, pch=21, bg=1, cex=1, xaxt='n', ylab="Probability of resistance", cex.lab=1.5, cex.axis=1.65)
points(seq(0,5,0.1)+0.25, emergeProb2, pch=21, bg=2)
axis(1,tick=TRUE,labels=FALSE)
mtext(side=1, line=1.5, "Drug dose", cex=1.65)
legend(x='topright', c("High mutation rate", "Low mutation rate"), pch=21, pt.bg=c(1,2))

```

Obviously, this is a highly contrived situation.
A more interesting scenario would be when the ingestion rate **depends** on the drug dose.
If ingestion increases with drug dose, what happens to the probability of resistance emergence?
I will again compare at high and low mutation rates, allowing ingestion increase linearly from 7.5 (low enough to be unsustainable for the resistant mutant) at a drug dose of 0 to 10 at a drug dose of 10 (Fig. \@ref(fig:variableIngestFig)).

```{r, echo=FALSE, eval=FALSE}
lapply(seq(0,5,0.1), 
       function(drug) {
         print(drug)
         pars <- c(theta=7.5+0.5*drug, chi=0.1, alpha1=0.1, alpha2=0.09, eta1=1500, eta2=1500, epsilon1=20, epsilon2=20, delta1=0.1, delta2=0.1, mu=0.001, D=drug, tDrug=150)
         mclapply(seeds, 
                  function(s) stochastic_strain_competition_w_drugs(500, tstep=1, pars=pars, y0=c(R=unname(pars["theta"]/pars["chi"]), P1=10, P2=0), seed=s),
                  mc.cores=4)
       }) -> out_drug_variable_anorexia_hi

lapply(seq(0,5,0.1), 
       function(drug) {
         print(drug)
         pars <- c(theta=7.5+0.5*drug, chi=0.1, alpha1=0.1, alpha2=0.09, eta1=1500, eta2=1500, epsilon1=20, epsilon2=20, delta1=0.1, delta2=0.1, mu=0.0001, D=drug, tDrug=150)
         mclapply(seeds, 
                  function(s) stochastic_strain_competition_w_drugs(500, tstep=1, pars=pars, y0=c(R=unname(pars["theta"]/pars["chi"]), P1=10, P2=0), seed=s),
                  mc.cores=4)
       }) -> out_drug_variable_anorexia_lo
saveRDS(out_drug_variable_anorexia_hi, file="out_drug_variable_anorexia_hi.RDS")
saveRDS(out_drug_variable_anorexia_lo, file="out_drug_variable_anorexia_lo.RDS")


```

```{r variableIngestFig, echo=FALSE, fig.height=4, fig.width=6, fig.cap="Probability of drug-resistance emerging when ingestion depends on the drug dose."}

out_drug_anorexia_hi <- readRDS("out_drug_variable_anorexia_hi.RDS")
out_drug_anorexia_lo <- readRDS("out_drug_variable_anorexia_lo.RDS")

lapply(1:length(out_drug_anorexia_hi),
       function(i) (lapply(out_drug_anorexia_hi[[i]], function(o) (ifelse((!is.na(tail(o[,4],1))),tail(o[,4],10) %>% mean,0) > 20)) %>% unlist %>% sum)/length(out_drug_anorexia_hi[[i]])) %>% unlist -> emergeProb1

lapply(1:length(out_drug_anorexia_lo),
       function(i) (lapply(out_drug_anorexia_lo[[i]], function(o) (ifelse((!is.na(tail(o[,4],1))),tail(o[,4],10) %>% mean,0) > 20)) %>% unlist %>% sum)/length(out_drug_anorexia_lo[[i]])) %>% unlist -> emergeProb2

par(mar=c(3,5,0.5,0.5))
plot(seq(0,5,0.1), emergeProb1, pch=21, bg=1, cex=1, xaxt='n', ylab="Probability of resistance", cex.lab=1.5, cex.axis=1.65)
points(seq(0,5,0.1), emergeProb2, pch=21, bg=2)
axis(1,tick=TRUE,labels=FALSE)
mtext(side=1, line=1.5, "Drug dose", cex=1.65)
legend(x='topleft', c("High mutation", "Low mutation"), pch=21, pt.bg=c(1,2), bty='n')

```

