---
title: Stochastic simulation of evolution of antibiotic resistance in the face of
  anorexia
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

I am considering the following very simple model of resource-dependent competition between two parasite strains, $P_1$ and $P_2$:
\begin{align}
\frac{dR}{dt} &= \theta - \chi R - \frac{\alpha_1 R}{\eta_1 + R}P_1 - \frac{\alpha_2 R}{\eta_2 + R}P_2 \\
\frac{dP_1}{dt} &= \epsilon_1 \frac{\alpha_1 R}{\eta_1 + R}P_1 - \delta_1 P_1 - \mu P_1 + \mu P_2 \\
\frac{dP_1}{dt} &= \epsilon_2 \frac{\alpha_2 R}{\eta_2 + R}P_2 - \delta_2 P_2 - \mu P_2 + \mu P_1
\end{align}

In this model, resources come into the system through "ingestion," $\theta$, and are utilized by, or lost to, the host at the per-capita rate $\chi$. 
The two parasite strains each have a Type II functional response with separate acquisition rates and half-saturation constants, $\alpha_i$ and $\eta_i$. 
Each parasite converts resources into new parasites with efficiency $\epsilon_i$, and are killed at the rate $\delta_i$.
Note that we could approximate the effect of the immune system by changing the mortality rate from a linear function to a quadratic function (e.g., if you assume that the immune response dynamics are something like $dI/dt = a (P_1+P_2) - m I$, and you assume that this attains a quasi-equilibrium [e.g., a separation of timescales argument], then $\hat{I} = a (P_2+P_2)/m$ and the parasite mortality rate would be $\delta_1 P_1 I = \delta_1 a (P_1^2 + P_1 P_2)/m$).
Finally, there is mutation between the two strains at a rate $\mu$. 

The existence of mutation implies that a deterministic model will always have both strains present.
However, if you analyze a model where $\mu = 0$, then coexistence is impossible, and the species that will win is the one that suppresses resources the most.
The resource abundance at equilibrium for either strain alone is $\hat{R} = \eta_i \delta_i/(\epsilon_i \alpha_i - \delta_i)$, so if 
\begin{equation}
\frac{\eta_1 \delta_1}{\epsilon_1 \alpha_1 - \delta_1} > \frac{\eta_2 \delta_2}{\epsilon_2 \alpha_2 - \delta_2}
\end{equation}
then strain 1 will outcompete strain 2, and vice versa if the inequality is reversed.
 
Additionally, analysis shows that, in the absence of mutation, the single strain models always go to a stable equilibrium so long as
\begin{equation}
\frac{\theta}{\chi} > \frac{\eta_i \delta_i}{(\epsilon_i \alpha_i - \delta_i)},
\end{equation}
where $\theta/\chi$ is the resource equilibrium in the absence of infection. 
Thus, if hosts drop $\theta$ enough in the presence of infection, violating the above inequality, then no parasite strains can persist at all.

As a preliminary analysis, I will do stochastic simulations of the above model assuming that parasite strain 1 can persist and outcompete strain 2 in the absence of mutation.

```{r sim1}
stochastic_strain_competition <- function(tmax, tstep, pars, y0, seed) {
  set.seed(seed)
  
  theta <- pars["theta"]
  chi <- pars["chi"]
  alpha1 <- pars["alpha1"]
  alpha2 <- pars["alpha2"]
  eta1 <- pars["eta1"]
  eta2 <- pars["eta2"]
  epsilon1 <- pars["epsilon1"]
  epsilon2 <- pars["epsilon2"]
  delta1 <- pars["delta1"]
  delta2 <- pars["delta2"]
  mu <- pars["mu"]
  
  ## initialize the time at t = 0
  t <- 0
  
  R <- y0["R"]
  P1 <- y0["P1"]
  P2 <- y0["P2"]
  
  ## initialize a dataframe to store the time and population size
  sim <- data.frame(time=seq(0,tmax,tstep), 
                    R=rep(0, length(seq(0,tmax,tstep))),
                    P1=rep(0, length(seq(0,tmax,tstep))),
                    P2=rep(0, length(seq(0,tmax,tstep))))
                    
  sim[1,2:4] <- c(R,P1,P2)
  
  ## start the algorithm
  while (t < tmax) {
    ## compute rates
    rates <- c(theta, ## add resource = 1
               chi*R, ## host resource use = 2
               alpha1*R/(eta1+R)*P1, ## strain 1 resource consumption = 3
               alpha2*R/(eta2+R)*P2, ## strain 2 resource consumption = 4
               epsilon1*alpha1*R/(eta1+R)*P1, ## strain 1 replication = 5
               epsilon2*alpha2*R/(eta2+R)*P2, ## strain 2 replication = 6
               delta1*P1, ## strain 1 death = 7
               delta2*P2, ## strain 2 death = 8
               mu*P1, ## strain 1 mutating to strain 2 = 9
               mu*P2) ## strain 2 mutating to strain 1 = 10
   
    ## what time does the event happen?
    dt <- rexp(1, rate=sum(rates))
    
    ## if these are not equal, don't record anything
    if(max(which(t >= seq(0,tmax,tstep)))!=max(which(t+dt > seq(0,tmax,tstep))))
      sim[max(which(t+dt > seq(0,tmax,tstep))), 2:4] <- c(R, P1, P2)
    
    ## update t 
    t <- t + dt
    
    ## "wheel of fortune"
    wheel <- cumsum(rates)/sum(rates)
    
    ## which event happens? Draw a random uniform to determine
    rand <- runif(1)
    event <- 1 + sum(rand > wheel)
    if (event==1) R <- R + 1
    else if (event%in%2:4) R <- R - 1
    else if (event==5) P1 <- P1 + 1
    else if (event==6) P2 <- P2 + 1
    else if (event==7) P1 <- P1 - 1
    else if (event==8) P2 <- P2 - 1
    else if (event==9) { P1 <- P1 -1; P2 <- P2 + 1}
    else {P1 <- P1 + 1; P2 <- P2 + 1}
      
  }
  return(sim)
}

## choose parameters appropriately so that strain 1 can persist and outcompete strain 2
pars <- c(theta=10, chi=0.1, alpha1=1, alpha2=0.9, eta1=1000, eta2=1000, epsilon1=20, epsilon2=20, delta1=1, delta2=1, mu=0.01)
## resource abundance in the absence of parasites
pars["theta"]/pars["chi"]
## resource abundance with strain 1
R1 <- pars["eta1"]*pars["delta1"]/(pars["epsilon1"]*pars["alpha1"]-pars["delta1"])
R1
## strain 1 outcompetes strain 2 if this is smaller than the above
R2 <- pars["eta2"]*pars["delta2"]/(pars["epsilon2"]*pars["alpha2"]-pars["delta2"])
R2
## strain 1 abundance at equilibrium in the absence of mutation
-(((pars["eta1"] + R1)*(pars["chi"]*R1 - pars["theta"]))/(pars["alpha1"]*R1))

## This appears to be a pretty good set of parameters
## Set the initial conditions
y0 <- c(R = 100, P1 = 10, P2 = 0)

## Deterministic version of the model
deterministic_strain_competition <- function(t, y0, pars) {
  theta <- pars["theta"]
  chi <- pars["chi"]
  alpha1 <- pars["alpha1"]
  alpha2 <- pars["alpha2"]
  eta1 <- pars["eta1"]
  eta2 <- pars["eta2"]
  epsilon1 <- pars["epsilon1"]
  epsilon2 <- pars["epsilon2"]
  delta1 <- pars["delta1"]
  delta2 <- pars["delta2"]
  mu <- pars["mu"]
  
  R <- y0["R"]
  P1 <- y0["P1"]
  P2 <- y0["P2"]
  b <- pars["b"]
  bs <- pars["bs"]
  d <- pars["d"]
  ds <- pars["ds"]
  
  dRdt <- theta - chi*R - alpha1*R/(eta1+R)*P1 - alpha1*R/(eta1+R)*P1
  dP1dt <- epsilon1*alpha1*R/(eta1+R)*P1 - delta1*P1 - mu*P1 + mu*P2
  dP2dt <- epsilon2*alpha2*R/(eta2+R)*P2 - delta2*P2 - mu*P2 + mu*P1
  
  return(list(c(dRdt,dP1dt,dP2dt)))
}

library(deSolve)
det_sim <- ode(y = y0, times=seq(0,100,0.1), func=deterministic_strain_competition, parms=pars)

## Do some stochastic simulations
set.seed(1234)
seeds <- floor(runif(12, 1, 1e5))
system.time(stochastic_strain_competition(200, tstep=0.1, pars=pars, y0=y0, seed=seeds[1]))

library(parallel)
mclapply(seeds,
         function(s) stochastic_strain_competition(100, tstep=0.1, pars=pars, y0=y0, seed=s),
         mc.cores=12) -> stoch_sims

par(mfrow=c(1,3))
for (j in 2:4) {
  plot.new()
  plot.window(xlim=c(0,200), ylim=c(0, lapply(stoch_sims, function(s) max(s[,j])) %>% unlist %>% max))
  axis(1); axis(2); box('plot')
  for (i in 1:length(stoch_sims)) lines(stoch_sims[[i]][,1], stoch_sims[[i]][,j], col=gray(0.5))
  lines(stoch_sims[[1]][,1], lapply(stoch_sims, function(s) s[,j]) %>% do.call(cbind.data.frame, .) %>% apply(., 1, mean), col=1, lwd=2)
}


```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
